---
title: "The origin of FoodCoop products"
author: "Nele, Anna and Miri"
date: "2 März 2018"
output:
  pdf_document: default
  fig_caption: true
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(RSQLite)
#library(readr)
library(data.table)
library(sf)
library(rgdal)
library(dplyr)
library(raster)
library(tidyr)
library(foodstorage)
library(leaflet)
library("geosphere")

```

# Introduction

- title: Transport routes of the products of a FoodCoop?

The consumption of food has been identified as one of the most energy demanding activities within a household [@vringer1995]. Historically, patterns of food consumption have changed from locally produced food to a diet where the use of fruits, vegetables or spices from far away regions became common [@carlsson1998]. Also the number of stakeholders involved in food production decreased. The majority of food consumed today passes through a complex, indirect network of a few large, centralized producers, processors, transporters and distributors [@hill2008].

The “buylocal” food movement tries to reverse this trend and argues that locally produced food is not only fresher, richer in vitamins and better tasting, but also better for the environment due to shorter distances of transportation and less energy usage during production and transport [@grebitus2013]. The shorter distances food travels, the less fossil fuels are required for transport and the fewer greenhouse gases are emitted [@hill2008].Another advantage of locally produced food is the improvement of the economic situation of (especially small scale) farmers as they are able to reap a greater profit by the removement of brokers from the distribution chain [@hill2008]. However, several studies have questioned the economic benefits of local foods by stating that buying locally unaware of economic laws and principles is promoting a future characterized by less food security [e.g. @desrochers2012]. Other authors identified some shortcomings that emerge from relying solely on economic logic - for example the focus on efficiency (making the pie bigger) rather than distribution (dividing up the pie fairly) [@scharber2016]. All in all the idea of buying food directly from farmers and thereby shortening the transportation distance seems convincing. Therefore, the popularity of forming consumer cooperatives (so called foodcoops) has grown rapidly since the 1970s [@marion1975]. 

A foodcoop is the combination of people and households for joint shopping. The idea behind this is to collect large quantities of food directly from the producers, making it worthwhile for farmers to drive into the middle of a city. The same principle can be applied to organic foods from wholesalers.As a result members of a foodcoop can usually shop at a lower price than buying in retail stores. Also packaging of products can be avoided by ordering food in bulk and refilling it to reusable packagings. The foodcoops usually work by means of unpaid distributed work and collective administration [@foodcoopedia].

Such foodcoop were founded in Freiburg as well resulting in five existing foodcoops at the moment. The foodcoop "Kornkammer" was founded in 2015 and comprises about 50 households with more or less 100 members. The members decide together what kind of products they want to order, communicate with the producers and store the deliveries in a room where all members can purchase small quantities of the groceries.

This study tries to investigate the transport distances of the products consumed by the foodcoop "Kornkammer" and to inform the foodcoop members about the routes their food travels. A web app was created for this purpose.

# Methods
Data acquisition, data processing and data visualisation was performed by Mirjam, Anna and Nele working closely together. The version control "git" and the platform "github" made this cooperation easy and productive. 

## data acquisition 

The foodcoop "Kornkammer" uses a tablet with an app to manage the inventory of products, to enter deliveries and to process the purchases of the members. The app stores all bookings related to these tasks in an SQLite database. 

We used the database to extract information on the products available in the foodcoop, as their suppliers and their turnovers. For the SQL queries and all other computing we used the R software version 3.4.3 [@RCoreTeam]. The SQL queries were done with the R package "RSQLite". Next, we extended the information stored in the database with the adresses of the suppliers. 

We decided to devide the suppliers into three categories: 

- farmers (translation in web app: Erzeuger): A supplier who grows grains/legumes/etc. and directly sells its products to the foodcoop. Beekeepers were also summarized in this category.
- manufacturers (translation in web app: Produzent): A supplier who produces products like detergent or bread spread which consist of many ingredients. 
- whole salers (Translation in web app: Zwischenhändler): A supplier who bought products from growers or even other intermediate agents. Assosiations which support cooperatives by purchasing olive oil, cashew nuts or spices directly from the cooperatives were also summarized in this category.

Following, we tried to find out the origin of the products sold by the whole salers. We classified the accuracy of the information we got into three categories: 

- 1 = origin exactly known (with adress)
- 2 = region or country of product origin known
- 3 = information about origin not available, therefore only a guess of origin

Products of farmers are categorised with the category 1.

## data processing
In a next step we used the function "geocode" from the package "ggmap" to convert the adresses of our suppliers and the information about the origin of the products from whole salers into coordinates. In the cases, that only the information about the country of a product's origin was available, the resulting geoposition calculated by the geocode function was located in the middle of the country.

Afterwards we used the function "spDists" from the package "sp" to calculate the distances from the foodcoop to the suppliers and, if a supplier was a whole saler, the distance from the supplier to the location of the product's origin. The function "spDist" uses the Great Circle distance method on the WGS84 ellipsoid. The sum of the distances gave the total distance a product traveled. For those products, where no information about the origin was available, no total distance could be calculated and the products are represented as NA (missing values) in the following statistics.

Regarding the inaccuracy about the real distance a product had traveled, the purpose of the calculated distances is to illustrate the regional or global origin. Nevertheless the actual driving distance can also be estimated by using the function "osrmRoute" from the package "osrm".

For every product a Spatial Layer that stores its transport Route was created. The layers of all products are stored in one Shapefile, that contains all informations of the products in its attributes-table.

## data visualisation
The products average turnover per Year is analysed regarding the total distance the product had been transported and the accuracy of its origin. Additionaly the total distance is grouped into intervals, that have a quardratic increase with increasing distance. Barplots of the average turnover for every distance-interval are build.

To geovisualise the data, the generated Shapefiles are used. 
The Shapefile that includes the transport Routes of all products is visualised using the package leaflet, a tool to create interactive maps.

Three different display options were used to visualise the transport Routes between two locations:
 - tha actual driving distance
 - the shortes distance on the sheroid (great-circle or orthodromic distance)
 - a straight line on the map

# Results

```{r message=FALSE, warning=FALSE, include=FALSE}
# producersInfoStraight <- readOGR("../../data/producersInfoStraight/", "producersInfoStraight")
# producersInfo <- readOGR("../../data/producersInfo/", "producersInfo")
producersRoutes <- readOGR("../../data/producersRoutes/", "producersRoutes")
producersExist <- readOGR("../../data/producersExist/", "producersExist")
Kornkammer <- readOGR("../../data/Kornkammer/", "Kornkammer")

# names(producersInfo) <- c('Produkte_App', 'Produkte_Zusammenfassung', 'Produktgruppe', 'Verpackungseinheit', 'Lieferant', 'Ort', 'EntfernungZwischenhaendler', 'Herkunftsgenauigkeit', 'Lieferantentyp', 'EntfernungKK', 'Gesamtentfernung', 'n', 'turnover2015', 'turnover2016', 'turnover2017', 'avg.turnover')
# producersInfo$avg.turnover <- as.numeric(producersInfo$avg.turnover)
# producersInfo$Gesamtentfernung <- as.numeric(producersInfo$Gesamtentfernung)
# producersInfo$turnover2017 <- as.numeric(producersInfo$turnover2017)
# producersInfo$Produktgruppe <- as.character(producersInfo$Produktgruppe)
# producersInfo$Produkte_Zusammenfassung <- as.character(producersInfo$Produkte_Zusammenfassung)
```


The foodcoop gets 118 products from 21 different suppliers. 12 of the suppliers (57%) were classified as farmers, 2 (10%) as manufacturers and 7 (33 %) as whole salers.

The accuracy of information available on the origin of the products varied with the transportation distance of the products to the foodcoop (see Fig. \ref{fig:accuracy}). All products where the origin was exactly known were located within a 600 km radius around the foodcoop. The farer distances a product was transported, the more difficult it became to find out it's origin. Interestingly, the turnover per year was decreasing with the transportation distance to the foodcoop. For example the 5 products with the highest average turnover per year (oats, spelt, oat drink, sunflower seeds and spelt drink) were transportet maximum 300 km to the foodcoop whereas the products with the lowest turnover per year were spices from Nepal with transportation distances of almost 7000 km.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Plot of the average turnover of every product against the total distance the product has been transported from its origin to the FoodCoop. The accuracy of the origin is described with 3 levels: 1 = origin exactly known (with adress), 2 = region or country of product origin known, 3 = information about origin not available, therefore only a guess of origin. For some products not even a guess was possible. These are not displayed in the plot.\\label{fig:accuracy}"}
totalDistances <- read.csv2("../../data/totalDistances.csv")
totalDistances$Herkunftsgenauigkeit[which(totalDistances$Lieferantentyp=="Erzeuger")] <- 1
totalDistances$Herkunftsgenauigkeit <- as.factor(totalDistances$Herkunftsgenauigkeit)
ggplot(totalDistances, aes(Gesamtentfernung, avg.turnover, color = Herkunftsgenauigkeit)) + 
  geom_point() +
  xlab("total Distance [km]") + ylab("average turnover per year [kg]") +
  guides(color=guide_legend(title="Accuracy of origin"))

```

The analysis of transportation distances of products consumed by the foodcoop summarized to distance classes and product groups showed a similar picture: most of the products conusmed by the foodcoop were transported bewteen 100 and 200 km. The product groups within this distance class were staple food and drinks/juice. The class with the highest transportation distances (6400-12800 km) comprises koffee, spices and bread spread like peanut butter or chocolate spread (see Fig. XY).

For 53 out of 118 products not even a guess of the origin was possible and the share of products where no transportation distance could be calculated is the third largest ...




shows the actual driving directions for all products from its last transfer point. 


```{r createDistanceCateogriyFunction, message=FALSE, warning=FALSE, include=FALSE}
createDistanceCategory <- function(totalDistances) {
  
  newtotalDistances <- mutate(totalDistances, "Kategorie" = NA)
  
  newtotalDistances$Kategorie[newtotalDistances$Gesamtentfernung <= 100] <- "0-100"
  newtotalDistances$Kategorie[newtotalDistances$Gesamtentfernung > 100 & newtotalDistances$Gesamtentfernung <= 200] <- "100-200"
  newtotalDistances$Kategorie[newtotalDistances$Gesamtentfernung > 200 & newtotalDistances$Gesamtentfernung <= 400] <- "200-400"
  newtotalDistances$Kategorie[newtotalDistances$Gesamtentfernung > 400 & newtotalDistances$Gesamtentfernung <= 800] <- "400-800"
  newtotalDistances$Kategorie[newtotalDistances$Gesamtentfernung > 800 & newtotalDistances$Gesamtentfernung <= 1600] <- "800-1600"
  newtotalDistances$Kategorie[newtotalDistances$Gesamtentfernung > 1600 & newtotalDistances$Gesamtentfernung <= 3200] <- "1600-3200"
  newtotalDistances$Kategorie[newtotalDistances$Gesamtentfernung > 3200 & newtotalDistances$Gesamtentfernung <= 6400] <- "3200-6400"
  newtotalDistances$Kategorie[newtotalDistances$Gesamtentfernung > 6400 & newtotalDistances$Gesamtentfernung <= 12800] <- "6400-12800"
  newtotalDistances$Kategorie[is.na(newtotalDistances$Gesamtentfernung) == TRUE] <- "NA"
  
  return(newtotalDistances)
}
```

For those products, that didn't grow at its last transfer point, the distance from its last transfer poit to its origin can be calculated, in the case that its origin is known.


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="The distance the products had been transported from its origin to the FoodCoop, colored by the product groups. The products were grouped into intervals of the distance. The size of the intervals increases quadraticaly with increasing distance."}
newtotalDistances <- createDistanceCategory(totalDistances)

positions <- c("0-100", "100-200", "200-400", "400-800", "800-1600", "1600-3200", "3200-6400", "6400-12800", "NA")

ggplot(newtotalDistances, aes(x = Kategorie, y = avg.turnover, fill = Produktgruppe)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(limits = positions)+
  labs(#title = "Consumed products and their transport distance", 
       y = "Turnover [kg]",
       x = "Distance [km]")  +
  guides(color=guide_legend(title="Product group")) + 
         theme(axis.text.x = element_text(size=10, angle=45, hjust = 1))

```


For every product that has ever been ordered by the FoodCoop, the transportation route to the last transfer point, can always be calculated. For a product that is ordered directly from the farmer, the last transfer point equals to its origin. Figure \ref{fig1} : 


```{r map1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="The map with the driving directions from the suppliers to the FoodCoop \\label{fig1}"}
pal <- colorFactor(c("darkgreen", "blue", "red"), domain = c(  "Erzeuger", "Produzent", "Zwischenhaendler"))
KKIcon <- iconList(
  Kornkammer = makeIcon("../../data/icon-2.png", "../../data/icon-2.png", 18, 18)
)

leaflet(producersExist) %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(radius = 3,
                   stroke = FALSE, fillOpacity = 0.8, color=pal(producersExist$Lfrntnt),
                   popup = producersExist$Liefrnt) %>%  #, clusterOptions = markerClusterOptions()
  addLegend("bottomright",
            pal = pal, values = ~producersExist$Lfrntnt,
            title = "Type of supplier",
            opacity = 1
  ) %>%
  addMarkers(Kornkammer,
             lng = coordinates(Kornkammer)[1], lat = coordinates(Kornkammer)[2],
             icon =  ~KKIcon[Kornkammer$Name], popup= "Kornkammer", 
             label = "FoodCoop", 
             labelOptions = labelOptions(noHide = T,  textOnly = F, direction = "left")) %>% 
  addPolylines(data = producersRoutes, color = "black", opacity = 1, weight = 2)

```

shows the actual driving directions for all products from its last transfer point. 

```{r generateGeoData, message=FALSE, warning=FALSE, include=FALSE, error=TRUE}
############
con <- dbConnect(SQLite(), "../../data/kornInfo.sqlite")

#dbListTables(con)
#dbListFields(con, "kornumsatz_origin")
#dbListFields(con, "productInfo")
#dbListFields(con, "producerAdress")

productInfo <- dbGetQuery(con, "SELECT * FROM productInfo")
producerAdress <- dbGetQuery(con, "SELECT * FROM producerAdress")
kornumsatz <- dbGetQuery(con, "SELECT * FROM kornumsatz_origin")
origin <- dbGetQuery(con, "SELECT * FROM productOrigin")
Kornkammer <- dbGetQuery(con, "SELECT * from AdresseKornkammer")

productOrigin <- origin

KUperYear <- kornumsatz_perYear(kornumsatz = kornumsatz, productInfo = productInfo)
KU <- KUperYear %>% 
  spread(Jahr, Umsatz) %>% 
  mutate(avg = mean(c(`2016`, `2017`), na.rm = T))
names(KU) <- c("Produkte_Zusammenfassung", "turnover2015", "turnover2016", "turnover2017", "avg.turnover")

originWithDistances <- SupplierDistance(origin, producerAdress)

totalDistances <- totalDistancesFun(origin = origin, producers = producerAdress, productInfo = productInfo)

## count occurance of every product in the table, to split the turnover of the product to the different occurances.
totalDistances <- totalDistances %>% 
  add_count(Produkte_Zusammenfassung) %>% 
  left_join(KU, by = "Produkte_Zusammenfassung") %>% 
  mutate(turnover2015 = turnover2015 / n) %>% 
  mutate(turnover2016 = turnover2016 / n) %>% 
  mutate(turnover2017 = turnover2017 / n) %>% 
  mutate(avg.turnover = avg.turnover / n) %>% 
  mutate(Herkunftsgenauigkeit = ifelse(Lieferantentyp == "Erzeuger", 1, Herkunftsgenauigkeit))
  
meanDists <- totalDistances %>% 
  group_by(Produktgruppe) %>% 
  summarise(avgDistance = mean(Gesamtentfernung, na.rm=T))

###############################################
## prepare data for the plot:
producerAdress$xCoord <- as.numeric(producerAdress$xCoord)
producerAdress$yCoord <- as.numeric(producerAdress$yCoord)
## we only want to plot the producers where xCoordinates are available:
producersExist <- producerAdress

warning( paste0("The producers " , paste0(producerAdress[which(is.na(producerAdress$xCoord)),"Lieferant"], collapse = ", "), " cannot be diplayed"))
## St georgener BAuer: Untermühlbachhof
##  Kaiserstühler Hof: Hof Humberg (google vom stühli)

## ändern: Stefan zu Stefan Chab Honig Imker (oder so)

# convert producers to spatialpointsdataframe
coordinates(producersExist) <- ~xCoord + yCoord

# create productOrigin SpatialPointsDataFrame only with existing origins:
productOriginExist <- productOrigin[!( is.na(productOrigin$xCoord) | is.na(productOrigin$yCoord)),]
coordinates(productOriginExist) <- ~xCoord + yCoord

coordinates(Kornkammer) <- ~xCoord + yCoord
crs(Kornkammer) <-  "+proj=merc +a=6378137 +b=6378137 +lat_ts=0.0 +lon_0=0.0 +x_0=0.0 +y_0=0 +k=1.0 +units=m +nadgrids=@null +wktext  +no_defs"

crs(producersExist) <- "+proj=merc +a=6378137 +b=6378137 +lat_ts=0.0 +lon_0=0.0 +x_0=0.0 +y_0=0 +k=1.0 +units=m +nadgrids=@null +wktext  +no_defs"

crs(productOriginExist) <- crs(producersExist)

#writeOGR(productOriginExist, "data/productOriginExist/", "productOriginExist", "ESRI Shapefile")
#writeOGR(producersExist, "data/producersExist/", "producersExist", "ESRI Shapefile")
#writeOGR(Kornkammer, "data/Kornkammer/", "Kornkammer", "ESRI Shapefile")


dbDisconnect(con)

##################

```


```{r, echo=FALSE, error=TRUE, message=FALSE, warning=FALSE}

##########
createCurves <- function(i, tD = totalDistances, pEx = producersExist, pOE = productOriginExist){
  productOriginExist <- pOE
  producersExist <- pEx
  totalDistances <- tD
  pE <- which(producersExist$Lieferant == totalDistances$Lieferant[i])
  if(totalDistances$Lieferantentyp[i] == "Zwischenhaendler" & 
     totalDistances$Lieferant[i] %in% unique(productOriginExist$Lieferant) & length(pE) > 0){
    
    z <- which(productOriginExist$Lieferant == totalDistances$Lieferant[i] &
                 productOriginExist$Produkte_Zusammenfassung == totalDistances$Produkte_Zusammenfassung[i] &
                 productOriginExist$Ort == totalDistances$Ort[i])
    Li <- Lines(list(Line(gcIntermediate(coordinates(Kornkammer), coordinates(producersExist)[pE,], addStartEnd = T)), 
                     Line(gcIntermediate(coordinates(producersExist)[pE,], coordinates(productOriginExist)[z,], addStartEnd = T))) , ID = i)
  } else {
    if(length(pE)>0){
      Li <- Lines(Line(gcIntermediate(coordinates(Kornkammer), coordinates(producersExist)[pE,], addStartEnd = T)), ID = i)
      } else {Li <- Lines(Line(coords = cbind(x = c(0,0), y = c(0,0))), ID = i)}
  }
  return(Li)
}


liste <- list()
for(i in 1:nrow(totalDistances)){
  liste[[i]] <- createCurves(i)
}

producersL <- SpatialLines(liste, proj4string = crs(producersExist)) #crs(producersExist)
producersInfo <- SpatialLinesDataFrame(producersL, totalDistances)

pal2 <- colorNumeric(
  palette = "viridis",
  domain = producersInfo$avg.turnover, n = 10, reverse = F)
dashs <- producersInfo$Herkunftsgenauigkeit

dashs[which(dashs == 2)] <- 1
dashs[which(dashs == 1)] <- 0
dashs[which(dashs != 0)] <- 10
dash <- as.character(dashs)

leaflet(producersInfo) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolylines(weight = ifelse((producersInfo$avg.turnover/15) < 2, 2, (producersInfo$avg.turnover)/15),
               color = ~pal2(producersInfo$avg.turnover),
               popup = producersInfo$Produkte_Zusammenfassung, dashArray = dashs) %>% 
  addLegend(pal = pal2, values = ~producersInfo$avg.turnover, title = "avg.turnover",
            labFormat = labelFormat(suffix = " kg/yr"))

```

# Discussion

- uncertainties of data acquisition
- missing data
- transportation of seeds and seedlings/manure/etc. not included
very few studies have analyzed transportation
upstream of the farm (e.g., transport of farm equipment and
supplies to the farm), which may be important for life-cycle
GHG emissions.

 ... no caclulation of the actual driving routes because of inaccuracy of the information - no importance ... especcially for products from far.

Thus, transportation may be one of the main targets when a conventional energy analysis
is used, but not when greenhouse gas emissions from all sources are accounted for. When
a conventional energy analysis is used for food, important emissions which affect the climate
are not included. This may lead to a sub-optimisation of pollution control.

There is some debate on the extent
to which buying products based solely on the criteria of miles trav-
eled is actually a good indicator of environmental impact, given that
many studies show that distance traveled is actually a poor indicator
of total environmental impact (e.g., see Coley et al., 2009; Mason
et al., 2002; Teisl, 2011).

The large amount of products, whose origin is unknown and therefore represented in the results as missing values (NA) is mostly caused due to the missing information of the wholesale "Bode". This gap of information could either be filled with data from a further research, or a change of the supplier could be considered. The latter is an option, since most products ordered from "Bode" are also available at other existing suppliers and for only few products a new supplier is to be found.

- further analytics

- calculation was done with turnover in kg, but the data is only in this unit for most products, some data has the unit liters or piece, where the the worst case is the chocolate, where the piece only weights 100 g and the data is wrong with a factor of 10.

- some products from suppliers come even from sub-suppliers, who get the products form its real origin! (e.g. some products from OBEG are from BYODO)

## Conclusion

## References



